Preliminary insights into the use of pottery and culinary practices at Guijiabao site in southwest China

### 2024-04-28 analysis/paper/paper.Rmd
minor edits

<!--Guijaibao as a research area-->
This study primarily focuses on the Guijiabao site located in the Hengduan Mountain ranges, a region in southwest China that connects the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south. As a crucial passage for population movement, the Hengduan Mountain ranges and valleys not only facilitate material exchange [@Tong1986; @Hein2016; -@Hein2014; @Ma2021], but also contribute to the spread of agriculture further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017]. Guijiabao is a key site with the so far earliest Neolithic human occupation in this region, dating to 5000 BP and continuing to the historical period about 500 BP. [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022]. The occupational history at Guijiabao provides a unique opportunity to explore questions related to subsistence since the Neolithic period in a mountainous area. For instance, what role did agriculture play in local environments? What other resources were obtained in addition to crops? In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site using organic residue analysis of lipids. For comparison, we also applied the same method to a sample from the Gaoshan site on the Chengdu Plain.

@@

<!--discussion about possible ruminant fat origin-->
Neolithic potsherds from Guijiabao and Gaoshan (ca. 5000-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1). The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from deer, goat/sheep, or cattle based on the faunal remains from the site [@Zhou2019] and archaeological contexts in northwestern China [@Sun2023], alongside a modern database [@Evershed2002]. At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2). A C~3~ diet is supported by the predominance of C~3~ weeds at Guijiabao, such as *Pennisetum* and *Chenopodiaceae* [@Guijiabao2016]. Some species in *Chenopodiaceae* can undergo C~4~ photosynthesis in arid environments [@Rudov2020; @Chasan2022], which may explain the mixture of plant types in SYG-TN13-E22-2#2. The direct input from C~4~ plants could be another reason for the mixed signal, as we have identified millet in the same sample. The husbandry of ruminant animals detected in the samples remains unclear. It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats. Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at Guijiabao involving hunting to a certain degree.

@@

<!--Wild non-ruminant as possible non-ruminant fat origin-->
The possibility of terrestrial animal input for the non-ruminant fat, such as wild boar, cannot be ruled out. The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022]. Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period, given the only sporadic occurrences in the south [@Zhang2022]. Domesticated horses probably did not appear in southwest China until the Dian culture period (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic Guijiabao. Ongoing research on faunal remains at Guijiabao will provide more details for further discussion.

<!--ruminants and crops-->
Our preliminary data reveals the predominance of ruminants fed on a C~3~ and C~3~/C~4~ diet in pottery vessels during the Neolithic period, suggesting a potentially more significant role of ruminants in local culinary practices than previously thought. Most studies have primarily focused on mixed farming of millet and rice as the main economy in southwest China after the introduction of agricultural in the Neolithic period [@Deng2019]. At Guijaibao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022]. Nitrogen and carbon isotope analysis of human bones also suggests a more C~4~ diet [@Lin2022]. Additionally, stone knives found at the site might be potentially be used for farming practices [@Zhou2019; @Hao2022; @Lin2022]. In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet provides direct evidence of millet processing at Guijiabao. Our non-ruminant specimen (SYG-TN13-E23-3), noted above, also indicates the presence of rice, though further analysis is needed for confirmation. It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining why only one specimen in our sample had a non-ruminant signal. Nevertheless, the frequent processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption. This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022], or to reduce the impact of potential crop failure [@Wang2023tibetan].

The two historical potsherds (ca. 1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period. In addition to rice, the source of C~3~ plants could also have included wheat, which was introduced into the region during the late Bronze Age (ca. 2,800–2000 cal. BP) [@Huan2022]. It is surprising that ruminant adipose fat, rather than non-ruminant fat, was detected in the Gaoshan potsherd, given the predominance of domestic pigs in the faunal assemblage at the site [@He2020]. One possibility might be our Gaoshan potsherd is an outlier, potentially not representative of the broader dietary patterns. Another plausible explanation could be that the isotopic values of ruminant adipose fat result from a mixture of non-ruminant adipose fats and ruminant milk fats. This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002]. While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels. As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the southwest China.

@@

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and for conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper. The research was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).

>>>>

<!--Guijiabao as a research area-->
This study primarily focuses on the Guijiabao site located in the Hengduan Mountain ranges, a region in southwest China that connects the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south. As a crucial passage for population movement, the Hengduan Mountain ranges and valleys not only facilitate material exchange [@Tong1986; @Hein2016; -@Hein2014; @Ma2021], but also contribute to the spread of agriculture further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017]. Guijiabao is a key site with the so far earliest Neolithic human occupation in this region, dating to 5000 BP and continuing to the historical period about 500 BP. [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022]. The occupational history at Guijiabao provides a unique opportunity to explore questions related to subsistence since the Neolithic period in a mountainous area. For instance, what role did agriculture play in local environments? What other resources were obtained in addition to crops? In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site using organic residue analysis of lipids. For comparison, we also applied the same method to a sample from the Gaoshan site on the Chengdu Plain.

@@

<!--discussion about possible ruminant fat origin-->
Neolithic potsherds from Guijiabao and Gaoshan (ca. 5000-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1). The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from deer, goat/sheep, or cattle based on the faunal remains from the site [@Zhou2019] and archaeological contexts in northwestern China [@Sun2023], alongside a modern database [@Evershed2002]. At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2). A C~3~ diet is supported by the predominance of C~3~ weeds, such as *Pennisetum* and *Chenopodiaceae* [@Guijiabao2016]. Some species in *Chenopodiaceae* can undergo C~4~ photosynthesis in arid environments [@Rudov2020; @Chasan2022], which may explain the mixture of plant types in SYG-TN13-E22-2#2. The direct input from C~4~ plants could be another reason for the mixed signal, as we have identified millet in the same sample. The husbandry of ruminant animals detected in the samples remains unclear. It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats. Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at the site involving hunting to a certain degree.

@@

<!--Wild non-ruminant as possible non-ruminant fat origin-->
The possibility of terrestrial animal input for the non-ruminant fat, such as wild boar, cannot be ruled out. The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022]. Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period, given the only sporadic occurrences in the south [@Zhang2022]. Domesticated horses probably did not appear in southwest China until the Dian culture period (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic Guijiabao. Ongoing research on faunal remains from the site will provide more details for further discussion.

<!--ruminants and crops-->
Our preliminary data reveals the predominance of ruminants fed on a C~3~ and C~3~/C~4~ diet in pottery vessels during the Neolithic period, suggesting a potentially more significant role of ruminants in local culinary practices than previously thought. Most studies have primarily focused on mixed farming of millet and rice as the main economy in southwest China after the introduction of agricultural in the Neolithic period [@Deng2019]. At Guijiabao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022]. Nitrogen and carbon isotope analysis of human bones also suggests a more C~4~ diet [@Lin2022]. Additionally, stone knives found at the site might be potentially be used for farming practices [@Zhou2019; @Hao2022; @Lin2022]. In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet provides direct evidence of millet processing at Guijiabao. Our non-ruminant specimen (SYG-TN13-E23-3), noted above, might indicate the processing of rice, though further analysis is needed for confirmation. It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining why only one specimen in our sample had a non-ruminant signal. Nevertheless, the processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption. This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022], or to reduce the impact of potential crop failure [@Wang2023tibetan].

The two historical potsherds (ca. 1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period. In addition to rice, the source of C~3~ plants could also have included wheat, which was introduced into the region during the late Bronze Age (ca. 2,800–2000 cal. BP) [@Huan2022]. It is surprising that ruminant adipose fat, rather than non-ruminant fat, was detected in the Gaoshan potsherd, given the predominance of domestic pigs in the faunal assemblage at the site [@He2020]. One possibility might be our Gaoshan potsherd is an outlier, potentially not representative of the broader dietary patterns. Another plausible explanation could be that the isotopic values of ruminant adipose fat result from a mixture of non-ruminant adipose fats and ruminant milk fats. This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002]. 
Although determining changes in culinary practices over time and regional differences would require a larger sample size, our results contribute to improving our understanding of direct food practices during the spread of agriculture into southwest China through organic residues in ceramic vessels.

@@

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. Special thanks to Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for lipid extraction and analysis. We also thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. Thanks to Yin-Ru Chiang in the Biodiversity Research Center, and the Institute of Plant and Microbial Biology at Academia Sinica, Taiwan for their support during the experiment. We appreciate the valuable comments provided by the anonymous reviewers, which helped improve this paper. This work was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
